---
title: "CS201 Flexdashboard: Violent Crime In The US: Overview"
author: Faith Hardie
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    vertical_layout: scroll
    source_code: embed
    theme:  simplex
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(here)
library(janitor)
library(ggplot2)
library(scales)
library(plotly)
library(dplyr)
library(rnaturalearth)
library(rnaturalearthdata)
library(sf)
library(leaflet)
library(htmlwidgets)
library(usdata)

```

# Data

```{r include=FALSE}

data <- read_csv(here("data/report.csv"))

#data_2015 <- read_csv(here("data/data_2015.csv"))
#jurisdiction_homicide_data <- #read_csv(here("data/jurisdiction_homicide_data.csv"))
#jurisdiction_robbery_data <- #read_csv(here("data/jurisdiction_robbery_data.csv"))
```

```{r include=FALSE}
#cleaning up data 
data <- data %>%
  clean_names()

data$agency_code <- as.factor(data$agency_code)
data$agency_jurisdiction <- as.factor(data$agency_jurisdiction)
data$report_year<- as.factor(data$report_year)

```

## row

### Heat Map of US Violent Crime Rate 2015

```{r}
# Prepare crime data for 2015
data_2015 <- data %>%
  filter(report_year == 2015) %>%
  group_by(state = sub(".*, ", "", agency_jurisdiction)) %>%
  summarize(
    total_population = sum(population, na.rm = TRUE),
    total_violent_crimes = sum(violent_crimes, na.rm = TRUE),
    violent_crime_rate = (sum(violent_crimes, na.rm = TRUE) / sum(population, na.rm = TRUE)) * 100000
  ) %>%
  ungroup() %>%
  mutate(state = abbr2state(state)) # Convert abbreviations to full names

# Load U.S. state geometries
us_states <- ne_states(country = "United States of America", returnclass = "sf") %>%
  filter(!name %in% c("Hawaii", "Alaska")) # Exclude Hawaii and Alaska

# Merge crime data with spatial data
us_states <- us_states %>%
  left_join(data_2015, by = c("name" = "state"))

# Check for missing data
if (all(is.na(us_states$violent_crime_rate))) {
  stop("The violent_crime_rate column contains only NA values. Check the join or data preparation.")
}

# Create the heatmap with a title
map <- leaflet(us_states) %>%
  addTiles() %>%
  addPolygons(
    fillColor = ~colorNumeric(palette = "YlOrRd", domain = us_states$violent_crime_rate)(violent_crime_rate),
    weight = 1,
    color = "white",
    fillOpacity = 0.7,
    label = ~paste0(
      name, "<br>",
      "Violent Crime Rate: ", round(violent_crime_rate, 2), " per 100,000 people"
    )
  ) %>%
  addLegend(
    pal = colorNumeric(palette = "YlOrRd", domain = us_states$violent_crime_rate),
    values = ~violent_crime_rate,
    title = "Violent Crime Rate<br>(per 100,000)",
    position = "bottomright" # Move legend to top-left
  )

# Add a title to the map
map <- htmlwidgets::prependContent(
  map,
  htmltools::tags$h2("Violent Crime Rates US 2015", style = "text-align:center; margin: 10px;")
)

# Display the map
map
```

## row

### Robbery Trends 1975-2015

```{r}
# Aggregate 
jurisdiction_robbery_data <- data %>%
  group_by(jurisdiction = sub(",.*", "", agency_jurisdiction), report_year) %>%
  summarize(total_robberies = sum(robberies, na.rm = TRUE)) %>%
  ungroup()
##Creating smaller set: highestpopulation
top_jurisdiction <- jurisdiction_robbery_data %>%
  group_by(jurisdiction) %>%
  summarize(total_robberies_all_years = sum(total_robberies, na.rm = TRUE)) %>%
  arrange(desc(total_robberies_all_years)) %>%
  slice(1:10) %>%
  pull(jurisdiction)

# Filter data 
top_jurisdiction_data <- jurisdiction_robbery_data %>%
  filter(jurisdiction %in% top_jurisdiction)

##Need to mutate factor to numeric so that the x axis labels can change to every 5 years for readability purposes
top_jurisdiction_data <- top_jurisdiction_data %>%
  mutate(report_year = as.numeric(as.character(report_year)))  

# line chart
ggplot(top_jurisdiction_data, aes(x = report_year, y = total_robberies, color = jurisdiction, group = jurisdiction)) +
  geom_line(size = 1) +
  labs(
    title = "Robbery Trends for States with Highest Population",
    x = "Year",
    y = "Total Robberies",
    color = "Jurisdiction"
  ) +
   scale_x_continuous(
    breaks = seq(min(top_jurisdiction_data$report_year), max(top_jurisdiction_data$report_year), by = 5)
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.3),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  )
```

## row {.tabset .tabset-fade}

### US Homicides 1975 histogram

```{r}
### Aggregate homicides data
jurisdiction_homicide_data <- data %>%
  group_by(jurisdiction = sub(",.*", "", agency_jurisdiction), report_year) %>%
  summarize(total_homicides = sum(homicides, na.rm = TRUE)) %>%
  ungroup()

### Create smaller set: jurisdictions with the highest total homicides across all years
top_jurisdiction <- jurisdiction_homicide_data %>%
  group_by(jurisdiction) %>%
  summarize(total_homicides_all_years = sum(total_homicides, na.rm = TRUE)) %>%
  arrange(desc(total_homicides_all_years)) %>%
  slice(1:10) %>%
  pull(jurisdiction)

### Exclude "United States"
top_jurisdiction <- setdiff(top_jurisdiction, "United States")

### Filter data for top jurisdictions and year 1975
data_1975 <- jurisdiction_homicide_data %>%
  filter(jurisdiction %in% top_jurisdiction, report_year == 1975)

### Histogram for 1975
h1975 <- ggplot(data_1975, aes(x = jurisdiction, y = total_homicides, fill = jurisdiction)) +
  geom_col() +
  labs(
    title = "Total Homicides in 1975 for Top Jurisdictions",
    x = "Jurisdiction",
    y = "Total Homicides",
    fill = "Jurisdiction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10, angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplotly(h1975)
```

### US Homicides 2015 histogram

```{r}
### Aggregate homicides data
jurisdiction_homicide_data <- data %>%
  group_by(jurisdiction = sub(",.*", "", agency_jurisdiction), report_year) %>%
  summarize(total_homicides = sum(homicides, na.rm = TRUE)) %>%
  ungroup()

### Create smaller set: jurisdictions with the highest total homicides across all years
top_jurisdiction <- jurisdiction_homicide_data %>%
  group_by(jurisdiction) %>%
  summarize(total_homicides_all_years = sum(total_homicides, na.rm = TRUE)) %>%
  arrange(desc(total_homicides_all_years)) %>%
  slice(1:10) %>%
  pull(jurisdiction)

### Exclude "United States"
top_jurisdiction <- setdiff(top_jurisdiction, "United States")

### Filter data for top jurisdictions and year 2015
data_2015 <- jurisdiction_homicide_data %>%
  filter(jurisdiction %in% top_jurisdiction, report_year == 2015)

### Histogram for 2015
h2015 <- ggplot(data_2015, aes(x = jurisdiction, y = total_homicides, fill = jurisdiction)) +
  geom_col() +
  labs(
    title = "Total Homicides in 2015 for Top Jurisdictions",
    x = "Jurisdiction",
    y = "Total Homicides",
    fill = "Jurisdiction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, hjust = 0.5),
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    axis.text = element_text(size = 10, angle = 45, hjust = 1),
    legend.position = "none"
  )

ggplotly(h2015)

```

# About

This Dashboard was designed for my CS201 final project with the intent to share US crime statistics through readable charts.

The questions this dashboard explores are:

How has violent crime in the US changed since 1975-2015? How does violent crime vary from state to state?

The data set used for this dashboard is originally found at: <https://www.kaggle.com/datasets/marshallproject/crime-rates>

"Crime in Context, 1975-2015" The Marshall Project.

<<<<<<< HEAD
All Code for this project is available in my github: <https://github.com/FaithHardie/CS201_dashboard> 

All Code for this project is available in my github: <https://github.com/FaithHardie/CS201_dashboard>


